Two applications of logic programming to Coq

### 2020-11-06 papers/types2020/elab.tex
Minor adjustments.
Refactored sections a bit.  There is still a lot of inconsistency in the names of experts, etc.

%\newpage
\section{Elaboration of external proof certificates for the Coq
  kernel}

% Follow the text in \cite{blanco20coq}.

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic.
\begin{metanote}
  Small point: the Coq kernel can be soundly extended with axioms and will also validate classical tautologies
\end{metanote}
The rest of the Coq system, especially its
tactic language, is designed to help a human user build
proofs-as-$\lambda$-terms that can be checked by the kernel.

@@

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade}, we assume that the following step are taken.
%
\begin{enumerate}
\item Have an external prover output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  If one is interested in checking only one kind of external
proof structure, then the FPC for that structure could also be built
into the checker (via, say, partial evaluation of logic programs
\cite{lloyd91jlp}).

To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Bruijn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
as a proof of following theorem of Coq.
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
\end{lstlisting}

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

>>>>

\section{Elaboration of external proof certificates for the Coq
  kernel}

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic along with any axioms that have been adopted.
The rest of the Coq system, especially its tactic language, is
designed to help a human user build proofs-as-$\lambda$-terms that can
be checked by the kernel.

@@

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade,blanco20coq}, we assume that the following step are
taken.
%
\begin{enumerate}
\item Have an external prover output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  (We took exactly this step in Figure~\ref{fig:augmented}
where proof checking involved both proof certificates as well as
Coq-term structures.)  If one is interested in checking only one kind
of external proof structure, then the FPC for that structure could
also be built into the checker (via, say, partial evaluation of logic
programs \cite{lloyd91jlp}).

As an example, consider an external theorem prover for propositional
intuitionistic logic which return proof structures that are De Bruijn
expressions for the underlying untype $\lambda$-proof that encodes
(part of) the proofs it finds.  Consider the De Buijn expression
example at the end of Section~\ref{ssec:beyond}.  This can be used as
evidence that the following is a theorem of Coq.

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
\end{lstlisting}

Of course, the proof information in the untyped term must be
elaborated to be a fully typed and functional expression.  Proof
checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
of) the proof Figure~\ref{fig:kernel}.

### 2020-11-06 papers/types2020/sec3.tex
Refactored sections a bit.  There is still a lot of inconsistency in the names of experts, etc.

\newpage
\section{Proof theory and proof certificates}
\label{sec:three}

\newcommand{\XXi}{{\color{blue}{\Xi}}}
\newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}

\todo{Right now there is no intro to FPC}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

\subsection{Proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }

@@

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\trueExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\forall x:B.\, D} A}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodExpert{\XXi}{\XXi_1}{\XXi_2}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

\[
  \infer{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer{\Pscr\Downarrow A\vdash A}{}
  \qquad
  \infer{\Pscr\Downarrow\forall x. B(x)\supset D(x)\vdash A}
        {\vdash B(t)\qquad \Pscr\Downarrow D(t)\vdash A}
\]
Here, $\Pscr$ is a collection of Horn clauses, i.e., formulas of the
form $\forall \bar x_1. A_1\supset \forall \bar x_2. A_2\supset
\forall \bar x_n. A_n\supset A_0$.

@@

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment $\XXi\vdash t \colon A$ corresponds to a goal-reduction
step while the judgment $\XXi \Downarrow  l \colon D\vdash A$ corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

The proof system is parameterized by Coq' global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

@@

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauty of the Calculus of Inductive
Construction is that they are in fact defined inductively and
therefore the \emph{unfold} rule will handle those; this also
simplifies the syntax of proof-terms.

  % , and as is usual in Pure Type
% Systems we have the same grammar for terms and types
%
%, as shown in figure~\ref{fig:terms}.
% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
% %\begin{figure}
% \begin{align*}
% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
% l ::= & [] \;|\; (t\ ::\ l)
% \end{align*}
% % \caption{Grammar of terms}
% % \label{fig:terms}
% % \end{figure}

% we denote by $\Pscr_A$ the
% set of type declarations for the constructors of type $A$, of the form
% $k : D$ where $D$ is of the form
% $\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We
% denote by $\mathcal{A}$ the set of arity declaration for the inductive
% types.

%
%
%done Explain relations with other term assignments. Note that we do not use
% explicit substitutions (we are cut free)
% Explain the choice of not having certificates represent proof terms.
% \todo{  Explain that the unfold clause handles
% defined atoms and defined connectives (conjunction, disjunction, existential
% and also equality): this may be less efficient and proof-theoretically induce
% pointless phase changes, but it is arguably very elegant e simplifies proof
% terms that are inherited by Coq }

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of the signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.                                   % reification of Coq terms
  kind gref    type                                 % reification of refernces to global terms
  type global gref -> term.                      % coercion to term
  type indt   inductive -> gref.                 % reification of inductive types
  type indc   constructor -> gref.             % reification of their constructors 
  type app    list term -> term.               % reification of nary application
  type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}

@@

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC, for example iterative-deepening. Furthermore, the FPC can
provide a \emph{trace} that may be more customable that the one
offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\section{Proof theory and proof certificates}
\label{sec:three}

%\todo{Right now there is no intro to FPC}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

\subsection{Proof for the Horn fragment}
\label{ssec:focused}

@@

A \emph{Horn clause} is a formula of the form $\forall \bar
x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
atomic formulas.  It is well known that the following set of sequent
calculus proof rules are complete for both classical and
intuitionistic logic when one is attempting to prove that a given
atomic formula $A$ is provable from a give set $\Pscr$ of Horn
clauses. 
\[
  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
\]
Here, we use two different styles of sequents.  The sequent
$\Pscr\vdash A$ is the usual sequent which we generally use as the end
sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
\emph{focus} of the sequent.  The two left introduction rules and the
initial rule can only be applied to the focused formula.  This latter
point is in constrast to Gentzen's sequent calculus where these rules
can involve \emph{any} formula on the left of the $\vdash$.  The fact
that this proof system is complete for both classical and
intuitionistic logic (when restricted to the Horn clause fragment)
follows from rather simple considerations of Horn clauses
\cite{nadathur90jacm} and from the completeness of LJT proofs
\cite{Herbelin94}.  The use of the term \emph{focus} comes from
Andreoli's proof system for linear logic \cite{andreoli92jlc}.

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

@@

Figure~\ref{fig:augmented} contains these proof rules and extends it
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript).
  \item The inference rules are also anotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:kernel} corresponds to
the Calculus of Inductive Constructions restricted to the Horn clause
fragment of pure \lsti{Inductive} definitions.
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories. In our case, where proof
search steps are viewed as computation, the term judgment $\XXi\vdash
t \colon A$ corresponds to a goal-reduction step while the judgment
$\XXi \Downarrow l \colon D\vdash A$ corresponds to a backchaining
step. This coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelity of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

@@

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauty of the Calculus of Inductive
Construction is that they are, in fact,  defined inductively and
therefore the \emph{unfold} rule will handle those; this also
simplifies the syntax of proof-terms.

\subsection{Proof certificate checking}

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of the signatures of the notions that we use:
\begin{lstlisting}
kind term   type.                                   % reification of Coq terms
kind gref   type                                    % reification of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}

@@

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC, for example iterative-deepening. Furthermore, the FPC can
provide a \emph{trace} that may be more customable that the one
offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.

\subsection{Moving beyond Horn clauses}
\label{ssec:beyond}

If we do not restrict our attention to only Horn clauses then there is
no problem will presenting a focused proof system for all of
first-order intuitionistic logic using the connectives $\forall$ and
$\supset$.  The following rules extend those given earlier.
\[
  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As before, the syntactic variable $A$ denotes atomic formulas and, as
usual, the $\forall R$ rule has the restriction that $y$ is not free in
its conclusion.

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  As a result, an indexing mechanism
is invoked so that it is possible for the decide expert to properly
select locally added assumptions instead of just those assumptions
accepted globally.

*** To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Bruijn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
